---
title: "Regression Model Visualizations"
author: "Fateema Bazzi"
date: "2025-04-20"
output: 
  html_document:
    fig_width: 10
    fig_height: 8
params:
  metricsdir: "./"
  metadir: "./"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in Libraries
```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(gplots)
library(readxl)
library(ggrepel)
library(stringr)
library(ggstatsplot)
library(tibble)
library(rgl)
options(rgl.useNULL = TRUE)
library(gt)
library(ggpubr)
```

## Get directories from User
```{r, echo=FALSE}
metricsdir <- params$metricsdir
metadir <- params$metadir
print(metricsdir)
print(metadir)
```

## Load in Data
```{r, echo=FALSE}
all_metrics <- read.csv(paste(metricsdir, "all_regression_metrics.csv", sep=""))

metadata <- read.csv(paste(metadir,"metadata.csv", sep = ""))
row.names(metadata)<-metadata$drug

print(head(all_metrics))
print(head(metadata))
```

## Data Formatting
```{r, echo=FALSE}
#add a log 10 version of metrics for graphics
all_metrics$logMSE<- log10(all_metrics$MSE)
all_metrics$logRMSE<- log10(all_metrics$RMSE)

#Separate into model types
linear_metrics<-all_metrics[all_metrics$Model=="linear",]
row.names(linear_metrics)<- linear_metrics$Drug

elastnet_metrics<-all_metrics[all_metrics$Model=="elastic net",]
row.names(elastnet_metrics)<- elastnet_metrics$Drug

print(head(linear_metrics))
print(head(elastnet_metrics))
```

## Linear versus Elastic Net
### MSE
```{r, echo=FALSE}
#MSE
#wilcoxon signed-rank test
mse_sig_test<-wilcox.test(linear_metrics$MSE, elastnet_metrics$MSE, paired = TRUE)
mse_pval <- mse_sig_test$p.value

#Violin plot
ggbetweenstats(data = all_metrics, x = Model, y = logMSE,
               ylab = "log10(MSE)",
               bf.message = F, p.adjust.method = "none", 
               pairwise.display = "none",
               centrality.type = "nonparametric",
               centrality.label.args = list(size  = 6))+
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=20))
```

### RMSE
```{r, echo=FALSE}
#RMSE
#wilcoxon signed-rank test
rmse_sig_test<-wilcox.test(linear_metrics$RMSE, elastnet_metrics$RMSE, paired = TRUE)
rmse_pval <- rmse_sig_test$p.value

#violin plot
ggbetweenstats(data = all_metrics, x = Model, y = logRMSE,
               ylab = "log10(RMSE)",
               bf.message = F, p.adjust.method = "none", 
               pairwise.display = "none",
               centrality.type = "nonparametric",
               centrality.label.args = list(size  = 6))+
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=20))

```

### Pearson Correlation
```{r, echo=FALSE}
#pcorr
#wilcoxon signed-rank test
pcorr_sig_test<-wilcox.test(linear_metrics$Pearson.Correlation,
                            elastnet_metrics$Pearson.Correlation, 
                            paired = TRUE)
pcorr_pval <- pcorr_sig_test$p.value

#violin plot
ggbetweenstats(data = all_metrics, x = Model, y = Pearson.Correlation,
               ylab = "Pearson Correlation",
               bf.message = F, p.adjust.method = "none", 
               pairwise.display = "none",
               centrality.type = "nonparametric",
               centrality.label.args = list(size  = 6))+
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=20))
```

### Comparison table
```{r, echo=FALSE}
comp_tibble <- tibble(`Metric`=c("MSE","RMSE","Pearson Correlation"),
                      `Linear` = c(median(all_metrics[all_metrics$Model=="linear", "MSE"]), median(all_metrics[all_metrics$Model=="linear", "RMSE"]), median(all_metrics[all_metrics$Model=="linear", "Pearson.Correlation"])), 
                      `Elastic Net`= c(median(all_metrics[all_metrics$Model=="elastic net", "MSE"]), median(all_metrics[all_metrics$Model=="elastic net", "RMSE"]), median(all_metrics[all_metrics$Model=="elastic net", "Pearson.Correlation"])), 
                      `p-value` = c(mse_pval,rmse_pval, pcorr_pval))


comp_table<- gt(comp_tibble) |>
  tab_header(
    title = "Linear vs. Elastic Net Metrics",
  )

comp_table
```

## MSE Elastic Net
```{r, echo=FALSE}
#Sort by MSE
elastnet_metrics <- elastnet_metrics %>% arrange(MSE)
#get average and median
avg_MSE <- mean(elastnet_metrics$MSE)
med_MSE <- median(elastnet_metrics$MSE)

annotation <- data.frame(
   x = c(620),
   y = c(med_MSE+0.3),
   label = c(paste("Median=",round(med_MSE, digits = 3)))
)

#scatter plot
elastnet_metrics %>%
  ggplot( aes(x=Drug, y=MSE))+
    geom_point(color="red") +
    geom_hline(yintercept=med_MSE, linetype="dashed")+
    geom_label(data=annotation, aes( x=x, y=y, label=label),
               color="black",
               size=6, fontface="bold" )+
    theme_bw()+
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title = element_text(size = 20))

#histogram
elastnet_metrics %>%
  ggplot(aes(x=MSE))+
  geom_histogram(binwidth = 0.5, fill="#155289", color="#155289")+
  geom_vline(xintercept=med_MSE, linetype="dashed", color="red")+
  xlab("MSE")
```

Models with MSE < 1
```{r, echo=FALSE}
#models w/ mse below 1
table(elastnet_metrics$MSE<1)
length(elastnet_metrics$MSE[elastnet_metrics$MSE<1])/dim(elastnet_metrics)[1]
```
Models with MSE < 0.5
```{r,echo=FALSE}
#below 0.5
table(elastnet_metrics$MSE<0.5)
length(elastnet_metrics$MSE[elastnet_metrics$MSE<0.5])/dim(elastnet_metrics)[1]
```

### MSE Comparisons
```{r, echo=FALSE}
#MSE Comparisons
deepDR_MSE <- 1.98
mse_comp<- data.frame(
   model = c("Elastic Net", "DeepDR"),
   MSE = c(med_MSE, deepDR_MSE)
)
mse_comp %>%
ggplot(aes(x=model, y=MSE)) + 
  geom_bar(stat = "identity", fill="#155289", color="#155289")+
  theme_bw()+
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title = element_text(size=15),
          axis.text = element_text(size=15))

```

## RMSE Elastic Net
```{r, echo=FALSE}
#sort by RMSE values
elastnet_metrics <- elastnet_metrics %>% arrange(RMSE)
#get average and median
avg_RMSE <- mean(elastnet_metrics$RMSE)
med_RMSE <- median(elastnet_metrics$RMSE)

annotation <- data.frame(
   x = c(630),
   y = c(med_RMSE+0.1),
   label = c(paste("Median=",round(med_RMSE, digits = 3)))
)
#scatterplot
elastnet_metrics %>%
  ggplot( aes(x=Drug, y=RMSE)) +
    geom_point(color="red") +
    geom_hline(yintercept=med_RMSE, linetype="dashed")+
    geom_label(data=annotation, aes( x=x, y=y, label=label),
               color="black",
               size=6 , fontface="bold" )+
    theme_bw()+
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title = element_text(size = 20))

#histogram
elastnet_metrics %>%
  ggplot(aes(x=RMSE))+
  geom_histogram(binwidth = 0.1, fill="#155289", color="#155289")+
  geom_vline(xintercept=med_RMSE, linetype="dashed", color="red")+
  xlab("RMSE")
```

Models with RMSE < 1
```{r, echo=FALSE}
#models w/ rmse below 1
table(elastnet_metrics$RMSE<1)
length(elastnet_metrics$RMSE[elastnet_metrics$RMSE<1])/dim(elastnet_metrics)[1]
```
Models with RMSE < 0.5
```{r, echo=FALSE}
#below 0.5
table(elastnet_metrics$RMSE<0.5)
length(elastnet_metrics$RMSE[elastnet_metrics$RMSE<0.5])/dim(elastnet_metrics)[1]
```

### RMSE Comparisons
```{r, echo=FALSE}
deepTTA_RMSE <- 1.544
tCNNS_RMSE <- 2.017
BANDRP_RMSE <- 1.398
DeepCDR_RMSE <- 1.652

rmse_comp<- data.frame(
   model = c("Elastic Net", "DeepTTA", "tCNNS", "BANDRP", "DeepCDR"),
   RMSE = c(med_RMSE, deepTTA_RMSE, tCNNS_RMSE, BANDRP_RMSE, DeepCDR_RMSE)
)
rmse_comp %>%
ggplot(aes(x=model, y=RMSE)) + 
  geom_bar(stat = "identity", fill="#155289", color="#155289")+
  theme_bw()+
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title = element_text(size=15),
          axis.text = element_text(size=15))
```

## Pearson Correlation Elastic Net
```{r, echo=FALSE}
#sort by Pearson Correlation values
elastnet_metrics <- elastnet_metrics %>% arrange(Pearson.Correlation)
#get average and median
avg_pcorr <- mean(abs(elastnet_metrics$Pearson.Correlation))
med_pcorr <- median(elastnet_metrics$Pearson.Correlation)
annotation <- data.frame(
   x = c(630),
   y = c(med_pcorr+0.02),
   label = c(paste("Median=",round(med_pcorr, digits = 3)))
)
#scatter plot
elastnet_metrics %>%
    ggplot( aes(x=Drug, y=Pearson.Correlation, color=abs(Pearson.Correlation))) +
    geom_point(color="red") +
    geom_hline(yintercept=med_pcorr, linetype="dashed")+
    geom_label(data=annotation, aes( x=x, y=y, label=label),
               color="black",
               size=6 , fontface="bold" )+
    theme_bw()+
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position = "none",
          axis.title = element_text(size=20))

#histogram
elastnet_metrics %>%
  ggplot(aes(x=Pearson.Correlation))+
  geom_histogram(binwidth = 0.01, fill="#155289", color="#155289")+
  geom_vline(xintercept=med_pcorr, linetype="dashed", color="red")+
  xlab("Pearson Correlation")
```

Models with Pearson Correlation > 0.5
```{r, echo=FALSE}
#above 0.5
table(abs(elastnet_metrics$Pearson.Correlation)>0.5)
length(elastnet_metrics$Pearson.Correlation[abs(elastnet_metrics$Pearson.Correlation)>0.5])/dim(elastnet_metrics)[1]
```
Models with Pearson Correlation > 0.75
```{r, echo=FALSE}
#above 0.75
table(abs(elastnet_metrics$Pearson.Correlation)>0.75)
length(elastnet_metrics$Pearson.Correlation[abs(elastnet_metrics$Pearson.Correlation)>0.75])/dim(elastnet_metrics)[1]
```

### Pearson Correlation Comparisons
```{r, echo=FALSE}
deepTTA_pcorr <- 0.647
tCNNS_pcorr <- 0.371
BANDRP_pcorr <- 0.756
DeepCDR_pcorr <- 0.503

pcorr_comp<- data.frame(
   model = c("Elastic Net", "DeepTTA", "tCNNS", "BANDRP", "DeepCDR"),
   Pearson.Correlation = c(med_pcorr, deepTTA_pcorr, tCNNS_pcorr, BANDRP_pcorr, DeepCDR_pcorr)
)
pcorr_comp %>%
ggplot(aes(x=model, y=Pearson.Correlation)) + 
  geom_bar(stat = "identity", fill="#155289", color="#155289")+
  theme_bw()+
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title = element_text(size=15),
          axis.text = element_text(size=15))
```

## Comparisons all Metrics
```{r, echo=FALSE}
comp_tibble <- tibble(`Model`= c("Elastic Net","DeepTTA","DeepDR", "tCNNS", "BANDRP", "DeepCDR"),
                      `MSE` = c(round(med_MSE, digits = 3), "-", deepDR_MSE, "-", "-", "-"),  
                      `RMSE`= c(round(med_RMSE, digits = 3), deepTTA_RMSE, "-", tCNNS_RMSE, BANDRP_RMSE, DeepCDR_RMSE), 
                      `Pearson Correlation` = c(round(med_pcorr, digits = 3), deepTTA_pcorr, "[0.7, 0.96]", tCNNS_pcorr, BANDRP_pcorr, DeepCDR_pcorr))


comp_table<- gt(comp_tibble) |>
  tab_header(
    title = "Performance Comparison",
  )

comp_table
```

## Correlation of Metadata
```{r, echo=FALSE}
metadata$var_diff<- abs(metadata$train_ic50_var-metadata$test_ic50_var)

linear_metrics$total_cell_lines <- metadata$total_cell_lines[match(rownames(linear_metrics), rownames(metadata))]
linear_metrics$total_genes <- metadata$total_genes[match(rownames(linear_metrics), rownames(metadata))]
linear_metrics$train_ic50_var <- metadata$train_ic50_var[match(rownames(linear_metrics), rownames(metadata))]
linear_metrics$test_ic50_var <- metadata$test_ic50_var[match(rownames(linear_metrics), rownames(metadata))]
linear_metrics$var_diff <- metadata$var_diff[match(rownames(linear_metrics), rownames(metadata))]

elastnet_metrics$total_cell_lines <- metadata$total_cell_lines[match(rownames(elastnet_metrics), rownames(metadata))]
elastnet_metrics$total_genes <- metadata$total_genes[match(rownames(elastnet_metrics), rownames(metadata))]
elastnet_metrics$train_ic50_var <- metadata$train_ic50_var[match(rownames(elastnet_metrics), rownames(metadata))]
elastnet_metrics$test_ic50_var <- metadata$test_ic50_var[match(rownames(elastnet_metrics), rownames(metadata))]
elastnet_metrics$var_diff <- metadata$var_diff[match(rownames(elastnet_metrics), rownames(metadata))]
```

## Linear Regression Correlation
Test for normality
```{r, echo=FALSE}
shapiro.test(linear_metrics$MSE)
shapiro.test(linear_metrics$RMSE)
shapiro.test(linear_metrics$Pearson.Correlation)
shapiro.test(linear_metrics$total_cell_lines)
shapiro.test(linear_metrics$total_genes)
shapiro.test(linear_metrics$train_ic50_var)
shapiro.test(linear_metrics$test_ic50_var)
```
since none of the values follow a normal distribution, can't use Pearson correlation

### Correlation with Pearson Correlation
```{r, echo=FALSE}
ggscatter(linear_metrics, x = "total_cell_lines", y = "Pearson.Correlation",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Cell lines", ylab = "Pearson Correlation",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.x = 300, label.y = -0.5))

ggscatter(linear_metrics, x = "total_genes", y = "Pearson.Correlation", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Genes", ylab = "Pearson Correlation",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 0.6))

ggscatter(linear_metrics, x = "train_ic50_var", y = "Pearson.Correlation",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Training IC50 Variance", ylab = "Pearson Correlation",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.x = 7.5, label.y = -0.5))

ggscatter(linear_metrics, x = "test_ic50_var", y = "Pearson.Correlation", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Test IC50 Variance", ylab = "Pearson Correlation",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.x = 8, label.y = -0.5))
```

### Correlation with MSE
```{r, echo=FALSE}
ggscatter(linear_metrics, x = "total_cell_lines", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Cell lines", ylab = "MSE",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 1.3e+26))

ggscatter(linear_metrics, x = "total_genes", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Genes", ylab = "MSE",
          title = "Linear Regression Models")

ggscatter(linear_metrics, x = "train_ic50_var", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Training IC50 Variance", ylab = "MSE",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 1.3e+26))

ggscatter(linear_metrics, x = "test_ic50_var", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Test IC50 Variance", ylab = "MSE",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 1.3e+26))
```

### Correlation with RMSE
```{r, echo=FALSE}
ggscatter(linear_metrics, x = "total_cell_lines", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Cell lines", ylab = "RMSE",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 1.1e+13))

ggscatter(linear_metrics, x = "total_genes", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Genes", ylab = "RMSE",
          title = "Linear Regression Models")

ggscatter(linear_metrics, x = "train_ic50_var", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Training IC50 Variance", ylab = "RMSE",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 1.1e+13))

ggscatter(linear_metrics, x = "test_ic50_var", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Test IC50 Variance", ylab = "RMSE",
          title = "Linear Regression Models",
          cor.coeff.args = list(label.y = 1.1e+13))
```

## Elast Net Correlation
Test for normality
```{r, echo=FALSE}
shapiro.test(elastnet_metrics$MSE)
shapiro.test(elastnet_metrics$RMSE)
shapiro.test(elastnet_metrics$Pearson.Correlation)
shapiro.test(elastnet_metrics$total_cell_lines)
shapiro.test(elastnet_metrics$total_genes)
shapiro.test(elastnet_metrics$train_ic50_var)
shapiro.test(elastnet_metrics$test_ic50_var)
```
since none of the values follow a normal distribution, can't use Pearson correlation

### Correlation with Pearson Correlation
```{r, echo=FALSE}
ggscatter(elastnet_metrics, x = "total_cell_lines", y = "Pearson.Correlation", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Cell lines", ylab = "Pearson Correlation",
          title = "Elastic Net Models",
          cor.coeff.args = list(label.x = 300, label.y = -0.5))

ggscatter(elastnet_metrics, x = "total_genes", y = "Pearson.Correlation", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Genes", ylab = "Pearson Correlation",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "train_ic50_var", y = "Pearson.Correlation",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Training IC50 Variance", ylab = "Pearson Correlation",
          title = "Elastic Net Models",
          cor.coeff.args = list(label.x = 7, label.y = -0.1))

ggscatter(elastnet_metrics, x = "test_ic50_var", y = "Pearson.Correlation", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Test IC50 Variance", ylab = "Pearson Correlation",
          title = "Elastic Net Models",
          cor.coeff.args = list(label.x = 8, label.y = -0.1))
```

### Correlation with MSE
```{r, echo=FALSE}
ggscatter(elastnet_metrics, x = "total_cell_lines", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Cell lines", ylab = "MSE",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "total_genes", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Genes", ylab = "MSE",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "train_ic50_var", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Training IC50 Variance", ylab = "MSE",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "test_ic50_var", y = "MSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Test IC50 Variance", ylab = "MSE",
          title = "Elastic Net Models")
```

### Correlation with RMSE
```{r, echo=FALSE}
ggscatter(elastnet_metrics, x = "total_cell_lines", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Cell lines", ylab = "RMSE",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "total_genes", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Genes", ylab = "RMSE",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "train_ic50_var", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Training IC50 Variance", ylab = "RMSE",
          title = "Elastic Net Models")

ggscatter(elastnet_metrics, x = "test_ic50_var", y = "RMSE", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Test IC50 Variance", ylab = "RMSE",
          title = "Elastic Net Models")
```
